---
title: "DataCollection"

author: "Sam Ardery"

date: "`r Sys.Date()`"

output: 
  html_document: 
    toc: true 
    toc_float: true 
    theme: spacelab
    code_folding: hide
---

This document will collect all of the data that is needed for the expression comparison. The data is pulled from the directories containing the outputs of the gbrs pipeline for both paired-end(PE) and single-end(SE) sequencing files. 

# Packages
These packages are only loaded in this document. If you have not downloaded these packages, please do so before running this document.
```{r, message = FALSE}
library(assertthat)
library(tidyverse)
```


# Obtaining single-end data

This first chunk of code was adapted from code provided by Dr. Selcan Aydin and is used to isolate the gene expression counts from the SE analysis. 
```{r SE-directory-collection, message = FALSE, results = "hide"}
#Make total gene expression matrix from GBRS output
gbrsdir <- '/projects/munger-lab/SampleReruns/SE_Data'
gbrscountfiles <- list.files(gbrsdir, pattern=".diploid.genes.expected_read_counts",
  full=TRUE)
gbrstpmfiles <- list.files(gbrsdir, pattern=".diploid.genes.tpm",
  full=TRUE)
file1 <- read_tsv(gbrscountfiles[1])
read_gbrs <- function(filename) {
  samplename <- strsplit(basename(filename), "\\.")[[1]][1]
  str_replace(samplename, pattern = "_ES", replacement = "_")
  substring(samplename, 6) <- "."
  print(samplename)
  dat <- read_tsv(filename) %>% mutate(sample=samplename)
  dat$rep <- ifelse((grepl("_repB", filename) | grepl("_B", samplename)), "repB", "repA")
  dat <- unite(dat, id, sample, rep, remove=FALSE)
  assert_that(are_equal(dat$locus, file1$locus))
  select(dat, locus, id, sample, rep, total)
}
counts <- bind_rows(lapply(gbrscountfiles, read_gbrs)) %>%
  rename(count=total) 
tpm <- bind_rows(lapply(gbrstpmfiles, read_gbrs)) %>% 
  rename(tpm=total) 
```

This chunk will isolate the genes that have a nonzero count in more than 50% of the samples and a tpm greater than 0.5
```{r SE-significance-filter, message = FALSE}
# Take genes with tpm > 0.5 and >50% nonzero counts
SEfinal <- inner_join(counts, tpm) %>%
  group_by(locus) %>%
  filter(median(tpm) > 0.5, mean(count == 0) < 0.5) %>%
  select(-tpm)
print(paste("Gene Quantity: ", length(unique(SEfinal$locus))))    # ~15k genes
```

This chunk takes the count data from above and reformats it to allow for easy analysis and legibility. It also adds the biotype of each gene to the data table.
```{r SE-gene-by-sample-matrix}
#isolate the 3rd and fourth columns which contain the sample ids, the gene ids, and the counts
SE <- SEfinal[,-c(2,4)]
#round the count variable to four decimal places
SE$count <- format(round(SE$count, 4))
#reformat the table to make every sample a single column.
SE <- pivot_wider(SE, names_from = sample, values_from = count)
SE <- dplyr::rename(SE, GeneID = locus)
```


# Obtaining paired-end data

This is the same code with a different directory to create the same data for paired-end sequencing results.
```{r PE-directory-collection, message = FALSE, results = "hide"}
#Make total gene expression matrix from GBRS output
gbrsdir <- '/projects/munger-lab/SampleReruns/PE_Data'
gbrscountfiles <- list.files(gbrsdir, pattern=".diploid.genes.expected_read_counts",
  full=TRUE)
gbrstpmfiles <- list.files(gbrsdir, pattern=".diploid.genes.tpm",
  full=TRUE)
file1 <- read_tsv(gbrscountfiles[1])
read_gbrs <- function(filename) {
  samplename <- strsplit(basename(filename), "\\.")[[1]][1]
  str_replace(samplename, pattern = "_ES", replacement = "")
  substring(samplename, 6) <- "."
  print(samplename)
  dat <- read_tsv(filename) %>% mutate(sample=samplename)
  dat$rep <- ifelse((grepl("_repB", filename) | grepl("_B", samplename)), "repB", "repA")
  dat <- unite(dat, id, sample, rep, remove=FALSE)
  assert_that(are_equal(dat$locus, file1$locus))
  select(dat, locus, id, sample, rep, total)
}
counts <- bind_rows(lapply(gbrscountfiles, read_gbrs)) %>%
  rename(count=total) 
tpm <- bind_rows(lapply(gbrstpmfiles, read_gbrs)) %>% 
  rename(tpm=total)

```

This chunk will isolate the genes that have a nonzero count in more than 50% of the samples and a tpm greater than 0.5
```{r PE-significance-filter, message = FALSE}
# Take genes with tpm > 0.5 and >50% nonzero counts
PEfinal <- inner_join(counts, tpm) %>%
  group_by(locus) %>%
  filter(median(tpm) > 0.5, mean(count == 0) < 0.5) %>%
  select(-tpm)
print(paste("Gene Quantity: ", length(unique(PEfinal$locus))))    # ~15k genes
```

This chunk takes the count data from above and reformats it to allow for easy analysis and legibility. It also adds the biotype of each gene to the data table.
```{r PE-gene-by-sample-matrix}
#isolate the 3rd and fourth columns which contain the sample ids, the gene ids, and the counts
PE <- PEfinal[,-c(2,4)]
#round the count variable to four decimal places
PE$count <- format(round(PE$count, 4))
PE$sample <- str_replace(PE$sample, pattern = "_ES", replacement = "")
#reformat the table to make every sample a single column.
PE <- pivot_wider(PE, names_from = sample, values_from = count)
PE <- dplyr::rename(PE, GeneID = locus)
```

# Saving Data
This data is saved as an RData file to load into the expression analysis and any other document that requires the gene expression matrices. 
```{r Save-to-RData}
save(SE, PE, file = "DataCollection.RData")
save(PEfinal, file = "PEMappingData.RData")
save(SEfinal, file = "SEMappingData.RData")
```


